{\tt APFELgrid}: a high performance tool for parton density determinations

### 2016-05-06 paper.tex
minor edits

\begin{abstract}
  We present a new software package designed to reduce the computational
  burden of collider measurements in Parton Distribution Function (PDF)
  fits. The {\tt APFELgrid} package converts interpolated weight
  tables provided by {\tt APPLgrid} files into a more efficient format
  for PDF fitting by the combination with PDF and
  $\alpha_s$ evolution factors provided by {\tt APFEL}. This
  combination significantly reduces the number of operations required
  to perform the calculation of hadronic observables in PDF fits and
  simplifies the structure of the calculation into a readily optimised
  scalar product. We demonstrate that our technique can lead to a
  substantial speed improvement when compared to existing methods
  without any reduction in numerical accuracy.
  % We present a new method conceived to ease the inclusion of
  % hadronic observables into parton distribution function (PDF)
  % fits. Our technique relies on the the same principles adopted by
  % the existing fast interfaces but implements a set of improvements
  % aimed at optimizing the performance in the context of PDF fits
  % where many iterations and many predictions for each iteration are
  % usually required. The main novelties are the precomputation of the
  % PDF and $\alpha_s$ evolution and the optimisation of the numerical
  % convolution between hard cross sections and PDFs. We demonstrate
  % that our technique can lead to a substantial speed improvement as
  % compared to the existing interfaces.
\end{abstract}

@@

Measurements at colliders such as the Tevatron and Large Hadron Collider (LHC) have a unique capacity to shed light upon the internal dynamics 
of the proton and provide constraints upon proton PDFs~\cite{Rojo:2015acz}. However the inclusion of large hadron collider datasets 
into PDF fits can provide a significant challenge due to the large computational footprint of performing accurate
theoretical predictions over the many iterations required in a fitting procedure. In order to make the fullest use of current 
and future LHC results, efficient strategies for the computation of these observables must therefore be employed.

The typical Monte Carlo software packages used to perform predictions for 
hadron-collider observables cannot be easily deployed in a PDF fit due the processing time 
required to obtain accurate results (usually of the order of a few hours or more per data
point). To overcome such limitations, the typical strategy
adopted for fast cross section prediction relies on the precomputation
of the partonic hard cross sections in such a way that the standard numerical
convolution with any set of PDFs can be reliably approximated by means of
interpolation techniques.

@@

While these tools have proven to be invaluable in the extraction
of parton densities, the volume of experimental data made available by
LHC collaborations for use in PDF fits is already stretching the
capabilities of the typical fitting technology. A typical
global PDF fit may now include thousands of hadronic data points for which
predictions have to be computed thousands of times during the
minimisation process. As a consequence performing these predictions
using the standard interpolating tools, $i.e.$
{\tt APPLgrid} and {\tt FastNLO}, starts to become prohibitively time-consuming.
For this reason a high-performance tool tailored specifically to the requirements of PDF analysis becomes increasingly
important.

@@

A number of tools ($e.g.$ {\tt
  APFEL},
{\tt HOPPET}~\cite{Salam:2008qg} and {\tt QCDNUM}~\cite{Botje:2010ay}) 
are available which perform PDF evolution via an analogous
interpolation procedure. In such a way PDFs at a general scale $Q_\tau$ may be
expressed as a product of PDFs at some initial, fitting scale $Q_0$ and an
\textit{evolution operator} obtained by the solution of the DGLAP equation.
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where latin indices run over PDF flavour, greek indices run over points in an initial-scale interpolating $x$-grid and
the evolution operator $A$ may be accessed directly in the {\tt APFEL} package. Given this operator, we may replace the
(general-scale) PDFs used in the subprocess parton density
Eq.~(\ref{eq:APPLsubproc}) with their equivalent expressions evaluated at the fitting scale as
\begin{equation}\label{eq:FK1}
\begin{array}{rcl}
F^{(s)}_{\alpha\beta,\tau} &=&  \displaystyle \sum_{i,j} \sum_{k,l}
                               \sum_{\delta,\gamma} C^{(s)}_{ij}
                               \left[  A^\tau_{\alpha\delta ik}\;
                               f_k(x_\delta,Q^2_0) A^\tau_{\beta\gamma
                               jl}\; f_l(x_\gamma,Q^2_0) \right]\;\;\;
  \\
\\
&=& \displaystyle \sum_{k,l}\sum_{\delta,\gamma}
\widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,,
\end{array}
\end{equation}
with the object
\begin{equation}
  \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta} =
  \sum_{i,j} C^{(s)}_{ij} A^\tau_{\alpha\delta ik}
  A^\tau_{\beta\gamma jl}\,,
\end{equation}
combining the operations of subprocess density construction and PDF
evolution. Going further and using the expression for
subprocess parton densities in Eq.~(\ref{eq:FK1}) in the full
cross section calculation we obtain
\begin{equation}
\sigma_{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma}\sum_p
\sum_{s} \sum_{\alpha,\beta}
\sum_{\tau} 
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,.
\end{equation}
Performing some further contractions it is possible to obtain an
extremely compact expression for the calculation of the cross section
in question, in terms of only the initial-scale PDFs and summing only
over the initial scale interpolating $x$-grid and the incoming parton
flavours:
\begin{equation}\label{eq:FK}
  \sigma _{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma} 
  \widetilde{W}_{kl,\delta\gamma} \,f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0),
\end{equation}
where the object:
\begin{equation}\label{eq:FKTable}
  \widetilde{W}_{kl,\delta\gamma} = \sum_p\sum_{s}\sum_{\alpha,\beta} \sum_{\tau}
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)  W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
\end{equation}
is referred to here as an {\tt FK} table, and combines the information
stored in {\tt APPLgrid}-style interpolated weight grids with
analogously interpolated DGLAP evolution operators. This combination
enables for a maximally efficient expression for the
calculation of observables at hadron colliders under PDF variation, without invoking
any additional approximation.

@@

The {\tt FK} product of Eq.~(\ref{eq:FK}) differs with respect to the
product in Eq.~(\ref{eq:applconv}) in several notable ways. Firstly
the typical {\tt APPLgrid} or {\tt FastNLO} products use as input PDFs at a
general scale, requiring that PDF evolution {\it e.g.}
Eq.~(\ref{eq:fastPDFfinal_recalled}) be performed for every variation of the PDFs
during the fit. In the {\tt FK} product
this evolution is pre-cached at the stage of {\tt FK} table
generation, requiring only initial-scale PDFs at the time of fitting. 
This pre-caching of the evolution also removes the need to
 sum over hard scale and perturbative order during the
fit, further reducing the number of operations required. As the {\tt FK}
product acts directly at the fitting scale, it benefits from the typically reduced number of active
partonic modes, with the sum over flavours in
Eq.~(\ref{eq:FK}) being limited to those directly parametrised in the
fit. Having reduced the calculation to such a simple form, it is also
straightforward to apply standard 
computational tools such as multi-threading through e.g OpenMP or
Single Instruction Multiple Data (SIMD) operations e.g SSE or AVX to
further reduce computational expense.

@@

We shall now examine the performance differences between
the two expressions for fast interpolated cross section prediction
Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt
  FK}).
%
In order to provide a comprehensive benchmark, we consider here a wide range of
processes including LHC and Tevatron electroweak vector boson
production
measurements~\cite{Aaij:2012mda,Aaij:2012vn,Chatrchyan:2013mza,Chatrchyan:2013uja,Chatrchyan:2012xt,Aad:2013iua,Aad:2011fp,Aad:2011dm,Aaltonen:2010zza},
$t\bar{t}$ total
cross sections~\cite{ATLAS:2012aa, Chatrchyan:2013faa,Chatrchyan:2012bra,Chatrchyan:2012ria},
double-differential Drell-Yan
cross sections~\cite{Chatrchyan:2013tia,CMS:2014jea} and inclusive jet
data~\cite{Chatrchyan:2012bja,Aad:2011fc,Aad:2013lpa,Abazov:2007jy}. Predictions
are performed over a wide range of kinematics, for a total of 52
source {\tt APPLgrid} files corresponding to the majority of available
LHC and Tevatron datasets of applicability to PDF determination.
While the source {\tt APPLgrid} files have varying grid densities in
$x$ and $Q^2$, for the purposes of comparison the corresponding {\tt
  FK} tables are produced consistently with 30 points in $x$, and at
an initial scale below the charm threshold, therefore with seven
active partonic species. These settings are chosen so as to provide a realistic
comparison, in a production environment the density and distribution of the
$x$-grid may be adjusted to match interpolation accuracy requirements.  For these comparisons the {\tt FK}
table is stored as double-precision in memory for table generation and in
single-precision for the purposes of computing the product and memory overheads.

>>>>

\begin{abstract}
  We present a new software package designed to reduce the computational
  burden of hadron collider measurements in Parton Distribution Function (PDF)
  fits. The {\tt APFELgrid} package converts interpolated weight
  tables provided by {\tt APPLgrid} files into a more efficient format
  for PDF fitting by the combination with PDF and
  $\alpha_s$ evolution factors provided by {\tt APFEL}. This
  combination significantly reduces the number of operations required
  to perform the calculation of hadronic observables in PDF fits and
  simplifies the structure of the calculation into a readily optimised
  scalar product. We demonstrate that our technique can lead to a
  substantial speed improvement when compared to existing methods
  without any reduction in numerical accuracy.
  % We present a new method conceived to ease the inclusion of
  % hadronic observables into parton distribution function (PDF)
  % fits. Our technique relies on the the same principles adopted by
  % the existing fast interfaces but implements a set of improvements
  % aimed at optimizing the performance in the context of PDF fits
  % where many iterations and many predictions for each iteration are
  % usually required. The main novelties are the precomputation of the
  % PDF and $\alpha_s$ evolution and the optimisation of the numerical
  % convolution between hard cross sections and PDFs. We demonstrate
  % that our technique can lead to a substantial speed improvement as
  % compared to the existing interfaces.
\end{abstract}

@@

Measurements at colliders such as the Tevatron and Large Hadron Collider (LHC) have a unique capacity to shed light upon the internal dynamics 
of the proton and provide constraints upon proton PDFs~\cite{Rojo:2015acz}. However including large hadron collider datasets 
in PDF fits can provide a significant challenge due to the large computational footprint of performing accurate
theoretical predictions over the many iterations required in a fitting procedure. In order to make the fullest use of current 
and future LHC results, efficient strategies for the computation of these observables must therefore be employed.

The typical Monte Carlo software packages used to perform predictions for 
hadron collider observables cannot be easily deployed in a PDF fit due the processing time 
required to obtain accurate results (usually of the order of a few hours or more per data
point). To overcome such limitations, the typical strategy
adopted for fast cross section prediction relies on the precomputation
of the partonic hard cross sections in such a way that the standard numerical
convolution with any set of PDFs can be reliably approximated by means of
interpolation techniques.

@@

While these tools have proven to be invaluable in the extraction
of parton densities, the volume of experimental data made available by
LHC collaborations for use in PDF fits is already stretching the
capabilities of the typical fitting technology. A standard
global PDF fit may now include thousands of hadronic data points for which
predictions have to be computed thousands of times during the
minimisation process. As a consequence, performing these predictions
using the standard interpolating tools, $i.e.$
{\tt APPLgrid} and {\tt FastNLO}, starts to become prohibitively time-consuming.
For this reason a high-performance tool tailored specifically to the requirements of PDF analysis becomes increasingly
important.

@@

A number of tools ($e.g.$ {\tt
  APFEL},
{\tt HOPPET}~\cite{Salam:2008qg} and {\tt QCDNUM}~\cite{Botje:2010ay}) 
are available which perform PDF evolution via an analogous
interpolation procedure. In such a way PDFs at a general scale $Q_\tau$ may be
expressed as a product of PDFs at some initial fitting scale $Q_0$ and an
\textit{evolution operator} obtained by the solution of the DGLAP equation.
\begin{equation}\label{eq:fastPDFfinal_recalled}
  f_i(x_{\alpha},Q^2_\tau) = \sum_{k}
  \sum_\beta A^\tau_{\alpha\beta, ik}\;
  f_k(x_\beta,Q^2_0)\,, 
\end{equation}
where latin indices run over PDF flavour, greek indices run over points in an initial-scale interpolating $x$-grid and
the evolution operator $A$ may be accessed directly in the {\tt APFEL} package. Given this operator, we may replace the
(general-scale) PDFs used in the subprocess parton density
Eq.~(\ref{eq:APPLsubproc}) with their equivalent expressions evaluated at the fitting scale as
\begin{equation}\label{eq:FK1}
\begin{array}{rcl}
F^{(s)}_{\alpha\beta,\tau} &=&  \displaystyle \sum_{i,j} \sum_{k,l}
                               \sum_{\delta,\gamma} C^{(s)}_{ij}
                               \left[  A^\tau_{\alpha\delta ik}\;
                               f_k(x_\delta,Q^2_0) A^\tau_{\beta\gamma
                               jl}\; f_l(x_\gamma,Q^2_0) \right]\;\;\;
  \\
\\
&=& \displaystyle \sum_{k,l}\sum_{\delta,\gamma}
\widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,,
\end{array}
\end{equation}
with the object
\begin{equation}
  \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta} =
  \sum_{i,j} C^{(s)}_{ij} A^\tau_{\alpha\delta ik}
  A^\tau_{\beta\gamma jl}\,,
\end{equation}
combining the operations of subprocess density construction and PDF
evolution. Going further and using the expression for
subprocess parton densities in Eq.~(\ref{eq:FK1}) in the full
cross section calculation we obtain
\begin{equation}
\sigma_{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma}\sum_p
\sum_{s} \sum_{\alpha,\beta}
\sum_{\tau} 
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0)\,.
\end{equation}
Performing some further contractions it is possible to obtain an
extremely compact expression for the calculation of the cross section
in question, in terms of only the initial-scale PDFs and summing only
over the initial scale interpolating $x$-grid and the incoming parton
flavours:
\begin{equation}\label{eq:FK}
  \sigma _{pp\to X} = \sum_{k,l}\sum_{\delta,\gamma} 
  \widetilde{W}_{kl,\delta\gamma} \,f_k(x_\delta,Q^2_0) f_l(x_\gamma,Q^2_0),
\end{equation}
where the object:
\begin{equation}\label{eq:FKTable}
  \widetilde{W}_{kl,\delta\gamma} = \sum_p\sum_{s}\sum_{\alpha,\beta} \sum_{\tau}
\alpha_s^{p+p_{\rm LO}}(Q^2_\tau)  W_{\alpha\beta,\tau}^{(p)(s)} \widetilde{C}^{(s),\tau}_{kl,\alpha\beta\gamma\delta}
\end{equation}
is referred to here as an {\tt FK} table, and combines the information
stored in {\tt APPLgrid}-style interpolated weight grids with
analogously interpolated DGLAP evolution operators. This combination
enables for a maximally efficient expression for the
calculation of observables at hadron colliders under PDF variation, without invoking
any additional approximation.

@@

The {\tt FK} product of Eq.~(\ref{eq:FK}) differs with respect to the
product in Eq.~(\ref{eq:applconv}) in several notable ways. Firstly
the typical {\tt APPLgrid} or {\tt FastNLO} products use as input PDFs at a
general scale, requiring that PDF evolution {\it e.g.}
Eq.~(\ref{eq:fastPDFfinal_recalled}) be performed for every variation of the PDFs
during the fit. In the {\tt FK} product
this evolution is pre-cached at the stage of {\tt FK} table
generation, requiring only initial-scale PDFs at the time of fitting. 
This pre-caching of the evolution also removes the need to
 sum over hard scale and perturbative order during the
fit, further reducing the number of operations required. As the {\tt FK}
product acts directly at the fitting scale, it benefits from the typically reduced number of active
partonic modes, with the sum over flavours in
Eq.~(\ref{eq:FK}) being limited to those directly parametrised in the
fit. Having reduced the calculation to such a simple form, it is also
straightforward to apply standard 
computational tools such as multi-threading through e.g OpenMP or
Single Instruction Multiple Data (SIMD) operations such as SSE or AVX to
further reduce computational expense.

@@

We shall now examine the performance differences between
the two expressions for fast interpolated cross section prediction
Eq.~(\ref{eq:applconv}) ({\tt APPLgrid}) and Eq.~(\ref{eq:FK}) ({\tt
  FK}).
%
In order to provide a comprehensive benchmark, we consider here a wide range of
processes including LHC and Tevatron electroweak vector boson
production
measurements~\cite{Aaij:2012mda,Aaij:2012vn,Chatrchyan:2013mza,Chatrchyan:2013uja,Chatrchyan:2012xt,Aad:2013iua,Aad:2011fp,Aad:2011dm,Aaltonen:2010zza},
$t\bar{t}$ total
cross sections~\cite{ATLAS:2012aa, Chatrchyan:2013faa,Chatrchyan:2012bra,Chatrchyan:2012ria},
double-differential Drell-Yan
cross sections~\cite{Chatrchyan:2013tia,CMS:2014jea} and inclusive jet
data~\cite{Chatrchyan:2012bja,Aad:2011fc,Aad:2013lpa,Abazov:2007jy}. Predictions
are performed over a wide range of kinematics, for a total of 52
source {\tt APPLgrid} files corresponding to the majority of available
LHC and Tevatron datasets applicable to PDF determination.
While the source {\tt APPLgrid} files have varying grid densities in
$x$ and $Q^2$, for the purposes of comparison the corresponding {\tt
  FK} tables are produced consistently with 30 points in $x$, and at
an initial scale below the charm threshold, therefore with seven
active partonic species. These settings are chosen so as to provide a realistic
comparison, in a production environment the density and distribution of the
$x$-grid may be adjusted to match interpolation accuracy requirements.  For these comparisons the {\tt FK}
table is stored as double-precision in memory for table generation and in
single-precision for the purposes of computing the {\tt FK} product.
